---
title: "Chapter 12: Visualizing posteriors with bayesplot"
author: "Kjell Nygren"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chapter 12: Visualizing posteriors with bayesplot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(glmbayes)
library(bayesplot)
suppressPackageStartupMessages(library(ggplot2))
theme_set(bayesplot::theme_default())
```

## 1. Purpose

[**bayesplot**](https://mc-stan.org/bayesplot/) provides **ggplot2**-based plotting for Markov chains and Bayesian workflow diagnostics. Package **glmbayes** is built with **bayesplot** as a dependency and supplies an S3 **`pp_check()`** method for **`glmb`** objects (posterior predictive checks using **`simulate.glmb()`** outputs).

After fitting models in earlier chapters (**Chapters 07–08** overview; **Chapter 09** binomial illustration), visual summaries help assess mixing, posterior shape, and how well replicated data match the observed responses.

## 2. Minimal binomial fit (Menarche)

We reuse the **Menarche** grouped binomial illustration with default **`Prior_Setup()`** machinery.

```{r menarche-fit}
data(menarche, package = "MASS")

Age2 <- menarche$Age - 13
Menarche_Model_Data <- data.frame(
  Menarche  = menarche$Menarche,
  Total     = menarche$Total,
  Age2      = Age2
)

ps <- Prior_Setup(
  cbind(Menarche, Total - Menarche) ~ Age2,
  family = binomial(link = "logit"),
  data   = Menarche_Model_Data
)

fit_logit <- glmb(
  cbind(Menarche, Total - Menarche) ~ Age2,
  family  = binomial(link = "logit"),
  pfamily = dNormal(mu = ps$mu, Sigma = ps$Sigma),
  data    = Menarche_Model_Data,
  n             = 800,
  use_parallel  = FALSE
)
coef_draws <- as.matrix(fit_logit$coefficients)
```

## 3. Coefficient trace and density overlays

Because **`glmb()`** returns **i.i.d.** draws, you can think of the coefficient matrix as **one** posterior sample stream. **bayesplot** functions that require **multiple MCMC chains** (for example **`mcmc_dens_overlay()`**) are not appropriate here; **`mcmc_trace()`** and **`mcmc_dens()`** still provide useful **univariate** views of the simulated coefficients.

```{r mcmc-views, fig.height = 5}
bayesplot::mcmc_trace(coef_draws)
bayesplot::mcmc_dens(coef_draws)
```

## 4. Posterior predictive checks

For **binomial** models **`glmbayes`** works on **proportions** so **`y`** and **`yrep`** line up (**`?pp_check.glmb`**). In practice you usually call **`bayesplot::pp_check(fit, ...)`**; here we assemble **`y`** and **`yrep`** once so this vignette knits cleanly even alongside older installed builds (**Chapter 05** uses the **`pp_check()`** wrapper directly).

```{r ppc-setup}
## Observed success proportions (aligned with simulate.glmb for binomial)
y_obs <- Menarche_Model_Data$Menarche / Menarche_Model_Data$Total

pred_resp <- predict(fit_logit, type = "response")
nd <- max(1L, min(50L, nrow(pred_resp)))
pred_sub <- pred_resp[seq_len(nd), , drop = FALSE]

y_rep <- stats::simulate(
  fit_logit,
  pred = pred_sub,
  prior.weights = fit_logit$prior.weights
)

stopifnot(nrow(y_rep) == nd, ncol(y_rep) == length(y_obs))
```

```{r ppc, fig.height = 4}
bayesplot::ppc_hist(y = y_obs, yrep = y_rep)
```

Try other **`ppc_*`** helpers via the **`fun`** argument (see **`?bayesplot-helpers`**).


## See also

- **Chapter 08** — full **`glmb()`** walk-through and interpretation of printed summaries.  
- **`?simulate.glmb`**, **`?predict.glmb`**, **`?pp_check.glmb`** — mechanics behind replicated responses.
