---
title: "Core 11: Large models — GPU acceleration using OpenCL (backend installation)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Core 11: Large models — GPU acceleration using OpenCL (backend installation)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

GPU acceleration is an **optional** feature of **glmbayesCore**. Matrix-level
samplers --- `rglmb()`, `rlmb()`, and the envelope machinery they use --- run
fully on the CPU regardless of whether OpenCL is available. No setup is needed
for standard use.

Where GPU acceleration pays off is with **large models**: high-dimensional
predictor sets or large posterior sample sizes. The computationally intensive
work is envelope construction and evaluation --- gradient and log-posterior
calculations at each tangency grid point grow with model dimension and are
embarrassingly parallel. Passing `use_opencl = TRUE` can substantially reduce
wall time. See **`vignette("Core-08", package = "glmbayesCore")`** for what
is accelerated and why.

End users of the formula interface (`glmb()`, `lmb()` in **glmbayes**) call the
same backend; install and compile-time requirements below apply to
**glmbayesCore** (see also **`vignette("Chapter-16", package = "glmbayes")`**
for the **glmbayes**-centric checklist).

This chapter describes how to enable GPU acceleration for **glmbayesCore**.
Ensure **opencltools** (and **nmathopencl**) are OpenCL-ready, then install
**glmbayesCore** from source so its `configure` scripts can enable
`-DUSE_OPENCL`.

# What you see when you load glmbayesCore

When **glmbayesCore** is attached in an interactive session it checks whether
GPU acceleration appears feasible. If `glmbayesCore_has_opencl()` is already
`TRUE`, attach is silent.

If OpenCL was **not** compiled in **and** a GPU or OpenCL stack is detected on
the host, you may see:

```
Note: glmbayesCore provides full CPU capability in this session
(e.g. Prior_Setup(), rglmb(), rlmb()). GPU acceleration is recommended
for bigger models and appears available. Reinstall glmbayesCore from source
with OpenCL at compile time to enable it; see vignette("Core-11",
package = "glmbayesCore") for install instructions.
```

On a machine with no GPU and no OpenCL stack, attach is silent.

To suppress the message in scripts or automated workflows:

```r
options(glmbayes.quiet_opencl_startup = TRUE)
```

# Enabling GPU acceleration: three steps

Work through these steps in order. After each step you can check whether you
are done and skip the rest.

## Step 1: Check whether OpenCL is already enabled

```r
library(glmbayesCore)
glmbayesCore_has_opencl()
```

If this returns `TRUE`, pass `use_opencl = TRUE` to `rglmb()` or `rlmb()` and
you are done. Otherwise continue to Step 2.

## Step 2: Ensure dependencies are OpenCL-ready

**glmbayesCore** **Imports** **opencltools** and **nmathopencl**. Both must be
available; **opencltools** must be built with OpenCL support for a successful
GPU-enabled Core build.

Check **opencltools**:

```r
opencltools::has_opencl()
```

If this returns `FALSE`, follow
**`vignette("Chapter-01", package = "opencltools")`** to install GPU drivers,
OpenCL headers, and the ICD loader for your platform, then reinstall
**opencltools** from source.

Install **nmathopencl** from source when building **glmbayesCore** from source
(R-Universe hosts current builds):

```r
install.packages(
  c("opencltools", "nmathopencl"),
  repos = c("https://cloud.r-project.org", "https://knygren.r-universe.dev"),
  type = "source"
)
```

**What you need on your system** (brief summary; details in **opencltools**
Chapter 01):

| Component | What it provides | Needed for |
|-----------|-----------------|------------|
| GPU driver | Exposes hardware to the OS | Runtime |
| OpenCL headers (`CL/cl.h`) | Required at compile time | Source build |
| OpenCL ICD loader (`OpenCL.dll` / `libOpenCL.so`) | Dispatches to vendor runtime | Runtime |

All three must be present before Step 3.

## Step 3: Reinstall glmbayesCore from source

With the OpenCL environment confirmed, reinstall **glmbayesCore** from source.
The package `configure` / `configure.win` script detects OpenCL headers and
libraries and defines `-DUSE_OPENCL` when appropriate.

Prebuilt **CRAN** / **R-Universe** binaries of **glmbayesCore** are often
**CPU-only**. GPU support requires a **source** install after Step 2.

### Windows

Install **Rtools** (<https://cran.r-project.org/bin/windows/Rtools/>) and,
if needed, **devtools**:

```r
install.packages("devtools")
```

Then:

```r
install.packages(
  "glmbayesCore",
  repos = c("https://cloud.r-project.org", "https://knygren.r-universe.dev"),
  type = "source"
)
```

Or from GitHub:

```r
devtools::install_github("knygren/glmbayesCore")
```

### Linux / macOS

```r
install.packages(
  "glmbayesCore",
  repos = c("https://cloud.r-project.org", "https://knygren.r-universe.dev"),
  type = "source"
)
```

On macOS, Xcode Command Line Tools and a suitable compiler are required; see
**opencltools** Chapter 01 for platform notes.

### After the install

```r
library(glmbayesCore)
glmbayesCore_has_opencl()
#> [1] TRUE
```

# Verifying the setup

```{r, eval=FALSE}
diagnose_glmbayes()
```

The diagnostic combines **opencltools** host checks with compile-time status
for **glmbayesCore**. A clean report ends with OpenCL enabled for this package
(the printed label may still say "glmbayes" for historical reasons).

Common resolutions:

- **Driver not installed** → install or update your GPU vendor driver.
- **Headers not found** → install the OpenCL SDK; see **opencltools** Chapter 01.
- **Runtime not found** → install the ICD loader (`ocl-icd-libopencl1` on Linux; CUDA Toolkit on Windows).
- **Runtime probe failed** → run `clinfo` outside R; check `/etc/OpenCL/vendors/` on Linux.
- **Package not compiled with OpenCL** → repeat Step 2, then reinstall **glmbayesCore** from source.

# Running a GPU-accelerated model

Pass `use_opencl = TRUE` to **`rglmb()`** (or **`rlmb()`** for Gaussian models).
The **`Cleveland`** heart-disease data ship with **glmbayesCore**:

```{r, eval=FALSE}
library(glmbayesCore)
data("Cleveland")

form <- hd ~ age + sex + cp + trestbps + chol +
  fbs + restecg + thalach + exang + oldpeak + slope + ca + thal

ps <- Prior_Setup(form, family = binomial(logit), data = Cleveland)

y <- ps$y
if (is.factor(y)) y <- as.numeric(y) - 1L else y <- as.numeric(y)

t_cpu <- system.time({
  fit_cpu <- rglmb(
    n = 20000,
    y = y,
    x = as.matrix(ps$x),
    family = binomial(link = "logit"),
    pfamily = dNormal(mu = ps$mu, Sigma = ps$Sigma),
    Gridtype = 2,
    use_parallel = TRUE,
    use_opencl = FALSE,
    verbose = FALSE
  )
})

t_gpu <- system.time({
  fit_gpu <- rglmb(
    n = 20000,
    y = y,
    x = as.matrix(ps$x),
    family = binomial(link = "logit"),
    pfamily = dNormal(mu = ps$mu, Sigma = ps$Sigma),
    Gridtype = 2,
    use_parallel = TRUE,
    use_opencl = TRUE,
    verbose = FALSE
  )
})

t_cpu
t_gpu
summary(fit_gpu)
```

The GPU path should match the CPU posterior up to Monte Carlo variation; gains
are most visible for larger `n`, more predictors, and higher-dimensional grids.

# Appendix A: AMD GPUs on Linux (ROCm OpenCL)

AMD provides multiple OpenCL implementations on Linux, but only **ROCm OpenCL**
is fully supported and stable. On **Ubuntu 22.04 or 24.04 LTS**:

```sh
sudo apt-get install rocm-opencl-runtime
```

**Supported AMD GPUs** (ROCm) include RX 7900 / 7800 / 7700 series and Instinct
MI200 / MI300. Older Polaris / Vega / Navi 1x–2x GPUs are not supported by
ROCm. See **`vignette("Chapter-01", package = "opencltools")`** for full
verification steps.
