---
title: "Core 20-S03: From calibrated priors to pfamily objects — pfamily_list() and dGamma_list()"
author: "Kjell Nygren"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Core 20-S03: From calibrated priors to pfamily objects — pfamily_list() and dGamma_list()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_br <- requireNamespace("bayesrules", quietly = TRUE)
```

```{r setup}
library(glmbayesCore)
```

# Core 20-S03: From calibrated priors to `pfamily` objects

`Prior_Setup_GLMM()` (Core 20-S02) returns calibrated prior *settings*.
Samplers instead take `pfamily` objects, the same prior-family objects
`rglmb()` and `rlmb()` use (see `?pfamily`). Each one bundles a `prior_list`
with the simulation function that knows how to use it. Two functions do the
conversion:

- **`pfamily_list()`**: one `pfamily` per random coefficient, for the
  **Block 2** regressions of \(b_{\cdot k}\) on \(W_k\). This is the
  `pfamily_list` argument of `lmerb()` and `glmerb()`.
- **`dGamma_list()`**: one `dGamma()` per group, for the **Block 1**
  measurement variance when each group has its own \(\sigma^2_j\). This is
  the `dispersion_ranef` argument of `lmerb()`.

## 1. User interface

### 1.1 `pfamily_list()`

```r
pfamily_list(object, ptypes = NULL, ...)
print(x, components = NULL, ...)
```

`pfamily_list()` is an S3 generic with two methods:

| Method | One `pfamily` per | Allowed `ptypes` | Documented in |
|--------|-------------------|------------------|---------------|
| `pfamily_list.Prior_Setup_GLMM()` | random coefficient \(k\) | `"dNormal"` (default), `"dIndependent_Normal_Gamma"` | this vignette |
| `pfamily_list.Prior_SetupGroup()` | row group | `"dNormal"` (default), `"dNormal_Gamma"`, `"dIndependent_Normal_Gamma"` | Core 20-S04 |

`ptypes` can be a single string applied to every element, or one string per
element (named or in order). The result has class
`c("pfamily_list", "list")` and an attribute `"ptypes"` recording the
resolved choice. `print()` shows all elements by default. Use
`components =` (names or indices) to show a subset.

For a `Prior_Setup_GLMM` object, component \(k\) is built from
`pop.prior_list[[k]]` and `pop.ing_prior[[k]]`:

| `ptypes[k]` | Constructor call | Effect in the sampler |
|-------------|------------------|-----------------------|
| `"dNormal"` | `dNormal(mu, Sigma, dispersion = tau2_k)` | \(\tau^2_k\) fixed at the calibrated centre |
| `"dIndependent_Normal_Gamma"` | `dIndependent_Normal_Gamma(mu, Sigma, shape, rate, disp_lower, disp_upper)` | \(\tau^2_k\) also sampled, with a truncated Gamma prior on \(1/\tau^2_k\) |

The independent Normal–Gamma choice requires \(n_0 \le J\)
(`pop.dispersion.nprior` at most the number of groups, or equivalently
`pop.dispersion.pwt` at most 0.5).

### 1.2 `dGamma_list()`

```r
dGamma_list(object, ...)
## S3 method for class 'Prior_Setup_GLMM'
dGamma_list(object, max_disp_perc_measurement = NULL, ...)
print(x, groups = NULL, ...)
```

Requires a Gaussian `Prior_Setup_GLMM()` object fitted with
`dispformula = ~<group>`. Otherwise it stops with a message saying what to
change. Group \(j\) gets

```r
dGamma(shape = shape_ING_j, rate = rate_j, beta = <placeholder>,
       Inv_Dispersion = TRUE, max_disp_perc = mdp_j,
       disp_lower = lower_j, disp_upper = upper_j)
```

with values from `group.ing_prior[[j]]`. `max_disp_perc_measurement = NULL`
reuses each group's stored window. A scalar or a vector with one value per
group recomputes the windows without re-running `Prior_Setup_GLMM()`. The
result has class `c("dGamma_list", "list")` and three attributes:

| Attribute | Content |
|-----------|---------|
| `window_diagnostics` | Data frame, one row per group: `n_j`, `sigma2_hat`, `shape_ING`, `rate`, `max_disp_perc`, `disp_lower`, `disp_upper` |
| `measurement_prior_group` | The same `shape`/`rate`/`disp_lower`/`disp_upper` values as named vectors, in the form the per-group Gaussian engines take |
| `group.dispersion.fit`, `calibration_source` | Carried over from the setup so `lmerb()` can reuse the `glmmTMB` fit |

## 2. Example

Same data and formula as Core 20-S02, with one pooled-dispersion setup and
one per-group setup.

```{r data, eval = has_br}
data(big_word_club, package = "bayesrules")
dat <- big_word_club
dat$school_id <- factor(dat$school_id)
dat <- subset(
  dat,
  !is.na(score_ppvt) &
    !is.na(invalid_ppvt) & invalid_ppvt == 0L &
    complete.cases(dat[, c(
      "score_ppvt", "distracted_a1", "distracted_ppvt",
      "private_school", "title1", "free_reduced_lunch", "school_id"
    )])
)

form_lmer <- score_ppvt ~
  private_school + title1 + free_reduced_lunch +
  distracted_ppvt + distracted_a1 +
  free_reduced_lunch:distracted_a1 +
  (1 + distracted_ppvt + distracted_a1 || school_id)

design0 <- model_setup(form_lmer, data = dat, fit_mer = FALSE)
dat <- subset(dat, school_id %in% names(design0$groupef.rank)[design0$groupef.rank])
dat$school_id <- droplevels(dat$school_id)

ps_pooled <- Prior_Setup_GLMM(form_lmer, data = dat, pop.pwt = 0.01)

ps_group <- Prior_Setup_GLMM(
  form_lmer,
  data                 = dat,
  pop.pwt              = 0.01,
  dispformula          = ~school_id,
  group.max_disp_perc  = 0.8,
  group.dispersion.pwt = 0.1,
  group.alpha_target   = NULL
)
```

### 2.1 Normal priors with fixed \(\tau^2_k\) (default)

```{r pf-normal, eval = has_br}
pf <- pfamily_list(ps_pooled)
class(pf)
attr(pf, "ptypes")
print(pf, components = "distracted_a1")
```

Each element is an ordinary `pfamily`. Its `prior_list` holds the values the
sampler reads:

```{r pf-prior-list, eval = has_br}
str(pf[["distracted_a1"]]$prior_list)
```

### 2.2 Independent Normal–Gamma priors (sample \(\tau^2_k\))

```{r pf-ing, eval = has_br}
pf_ing <- pfamily_list(ps_pooled, ptypes = "dIndependent_Normal_Gamma")
attr(pf_ing, "ptypes")
str(pf_ing[["(Intercept)"]]$prior_list[c("shape", "rate", "disp_lower", "disp_upper")])
```

These are exactly `ps_pooled$pop.ing_prior[["(Intercept)"]]`.

### 2.3 Mixing prior types

Sample \(\tau^2\) for the random intercept only:

```{r pf-mixed, eval = has_br}
re <- names(ps_pooled$pop.prior_list)
ptypes_mixed <- stats::setNames(
  c("dIndependent_Normal_Gamma", rep("dNormal", length(re) - 1L)),
  re
)
attr(pfamily_list(ps_pooled, ptypes = ptypes_mixed), "ptypes")
```

### 2.4 Per-group measurement-variance priors

A pooled setup has no per-group calibration:

```{r dg-pooled-error, eval = has_br}
try(dGamma_list(ps_pooled))
```

With the per-group setup:

```{r dg, eval = has_br}
disp_pf <- dGamma_list(ps_group)
length(disp_pf)
print(disp_pf, groups = 1:2)
head(attr(disp_pf, "window_diagnostics"), 3L)
```

### 2.5 Changing the truncation windows

```{r dg-wide, eval = has_br}
disp_pf_wide <- dGamma_list(ps_group, max_disp_perc_measurement = 0.95)
cbind(
  narrow = head(attr(disp_pf,      "window_diagnostics")$disp_upper, 3L),
  wide   = head(attr(disp_pf_wide, "window_diagnostics")$disp_upper, 3L)
)
```

A wider window covers more of each \(\sigma^2_j\) posterior but usually
lowers the sampler's acceptance rate. The recomputed window uses the
**posterior** shape \(\text{shape}_{ING} + n_j/2\) with a rate matched to
`sigma2_hat`. The prior handed to the sampler (`shape_ING`, `rate`) does not
change.

### 2.6 Pooled measurement dispersion

With `dispformula = ~1` there is a single \(\sigma^2\). Its Gamma
calibration (`shape`, `rate`, `disp_lower`, `disp_upper`) is in
`ps_pooled$group.ing_prior`, so no list is needed.

### 2.7 Putting it together

```r
fit <- lmebayes::lmerb(
  form_lmer, data = dat,
  pfamily_list     = pfamily_list(ps_group),
  dispersion_ranef = dGamma_list(ps_group),
  dispformula      = ~school_id
)
```

`dispformula` must be the same in `Prior_Setup_GLMM()` and `lmerb()`. The
two arguments are independent, so the caller has to keep them consistent.

## 3. Lineage

### 3.1 Call trees

```
pfamily_list(object, ptypes)                    generic (R/pfamily_list.R)
 └─ pfamily_list.Prior_Setup_GLMM()             R/pfamily_list_Prior_Setup_GLMM.R
     ├─ validate / recycle / reorder ptypes     against names(pop.prior_list)
     ├─ n_prior_for() (local)                   pop.dispersion.nprior[k], or J w/(1-w)
     └─ for each RE coefficient k:
         ├─ "dNormal"
         │    └─ dNormal(mu, Sigma, dispersion)                      exported
         └─ "dIndependent_Normal_Gamma"
              ├─ stop if n_prior_k > J
              ├─ pop.ing_prior[[k]]   (or, if absent, rebuild with
              │                        .lmebayes_ing_limiting_posterior_window())
              └─ dIndependent_Normal_Gamma(mu, Sigma, shape, rate,
                                           disp_lower, disp_upper)   exported

print.pfamily_list(x, components)
 ├─ .pfamily_list_select_keys()                 names / indices → keys
 └─ print.pfamily() for each selected element

dGamma_list(object, ...)                        generic (R/dGamma_list.R)
 └─ dGamma_list.Prior_Setup_GLMM()              R/dGamma_list_Prior_Setup_GLMM.R
     ├─ stop unless gaussian and
     │   .lmebayes_ing_prior_is_grouped(group.ing_prior)
     ├─ .lmebayes_expand_scalar_or_vector()     max_disp_perc_measurement → per group
     ├─ for each group j:
     │   ├─ [window override] .lmebayes_ing_prior_quantile_window(
     │   │        shape_ING + n_j/2, sigma2_hat * (shape_ING + n_j/2 - 1), mdp_j)
     │   └─ dGamma(shape_ING, rate, Inv_Dispersion = TRUE, ...)      exported
     └─ attach window_diagnostics, measurement_prior_group,
        group.dispersion.fit, calibration_source

print.dGamma_list(x, groups)
 ├─ .lmebayes_select_named_list_keys()
 └─ print.pfamily() for each selected group
```

### 3.2 Helper reference

| Helper | File | Role |
|--------|------|------|
| `pfamily_list.Prior_Setup_GLMM()` | `pfamily_list_Prior_Setup_GLMM.R` | Block 2 method; one `pfamily` per random coefficient |
| `pfamily_list.Prior_SetupGroup()` | `Prior_SetupGroup.R` | Row-group method (Core 20-S04) |
| `.pfamily_list_select_keys()` | `pfamily_list.R` | Resolves `components =` for `print.pfamily_list()` |
| `dGamma_list.Prior_Setup_GLMM()` | `dGamma_list_Prior_Setup_GLMM.R` | Block 1 method; one `dGamma()` per group |
| `.lmebayes_select_named_list_keys()` | `dGamma_list.R` | Resolves `groups =` for `print.dGamma_list()` |
| `.lmebayes_ing_prior_is_grouped()` | `Prior_Setup_GLMM.R` | Tells the per-group and pooled `group.ing_prior` shapes apart |
| `.lmebayes_expand_scalar_or_vector()` | `mer_prior_helpers.R` | Recycles or reorders `max_disp_perc_measurement` to one value per group |
| `.lmebayes_ing_prior_quantile_window()` | `mer_prior_helpers.R` | Gamma-quantile window when a group's tail probability changes |
| `.lmebayes_ing_limiting_posterior_window()` | `mer_prior_helpers.R` | Fallback \(\tau^2_k\) window for setups without `pop.ing_prior` |
| `dNormal()`, `dIndependent_Normal_Gamma()`, `dGamma()` | `pfamily.R` | Exported constructors; see `?pfamily` and Core 02 |

### 3.3 Where the values come from

| `pfamily` field | Source in `Prior_Setup_GLMM()` (stage numbers from Core 20-S02) |
|-----------------|-------------------------------------------------------|
| Block 2 `mu`, `Sigma` | `pop.prior_list[[k]]` (Stage 7) |
| Block 2 `dispersion` (dNormal) | `pop.prior_list[[k]]$dispersion` (Stage 8) |
| Block 2 `shape`, `rate`, `disp_lower`, `disp_upper` | `pop.ing_prior[[k]]` (Stage 9) |
| Block 1 `shape`, `rate`, window | `group.ing_prior[[j]]` (Stages 10–11) |

## 4. Next

Core 20-S04 covers the row-group samplers that draw the Block 1 group
coefficients given Block 2.
