---
title: "Core 20-S01: Model setup and identifiability"
author: "Kjell Nygren"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Core 20-S01: Model setup and identifiability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_br <- requireNamespace("bayesrules", quietly = TRUE)
```

```{r setup}
library(glmbayesCore)
```

# Core 20-S01: Model setup and identifiability

`model_setup()` is the first step in every mixed-model workflow in
**glmbayesCore**. It takes an `lme4`-style formula and returns three things
in one object of class `"model_setup"`:

- the **sampler design**: the response `y`, the within-group design matrix
  `D`, the grouping factor `group`, and one group-level hyper-design matrix
  `W[[k]]` per random-effect coefficient;
- a **reference classical fit**: `lmer()` for Gaussian models, `glmer()`
  otherwise, plus an optional `glmmTMB` fit with per-group dispersion. These
  are used later to calibrate priors;
- **identifiability diagnostics** from `check_identifiability()` at both
  levels of the hierarchy.

Section 1 covers the user interface, Section 2 walks through an example and
Section 3 documents the internal lineage.

## 1. User interface

### 1.1 Formula rules

`model_setup()` accepts the formula language of `lme4::lmer()`, with three
structural rules:

1. **Exactly one grouping factor**, for example `school_id`.
2. **`||` for more than one random coefficient.** The sampler treats the
   random-effect covariance as diagonal, so write
   `(1 + x || school_id)` rather than `(1 + x | school_id)`. A random
   intercept alone must be written `(1 | school_id)`; `lme4` does not accept
   `(1 || school_id)`.
3. **Each fixed effect must be one of three types:**
   - a *group-level (level-2) covariate*: constant within each group, such
     as `private_school`;
   - a *population mean slope*: an observation-level covariate that also has
     a random slope, such as `distracted_ppvt`;
   - a *cross-level interaction* between a level-2 covariate and a random
     slope, such as `free_reduced_lunch:distracted_a1`.

Any other fixed term (for example an observation-level covariate with no
matching random slope) is rejected with an error naming the offending
columns.

These rules let `model_setup()` split the `lme4` fixed effects into one
group-level regression per random coefficient:

| Random coefficient \(k\) | Columns of `W[[k]]` |
|--------------------------|---------------------|
| `(Intercept)` | intercept plus **all** level-2 covariates |
| a random slope with a cross-level moderator | intercept (the population mean slope) plus the moderator |
| any other random slope | intercept only (the population mean slope) |

### 1.2 Arguments

| Argument | Default | Role |
|----------|---------|------|
| `formula` | — | Two-sided `lme4` formula obeying Section 1.1 |
| `data` | `NULL` | Data frame (falls back to the formula environment) |
| `family` | `gaussian()` | `gaussian()` uses `lmer()`; other families use `glmer()` |
| `REML` | `TRUE` | Passed to `lmer()` (and `glmmTMB`); ignored by `glmer()` |
| `control`, `start`, `verbose` | `NULL`, `NULL`, `0L` | Passed to the reference fit |
| `subset`, `weights`, `na.action`, `offset`, `contrasts` | — | Applied identically to design extraction and every reference fit, so they use the same rows |
| `devFunOnly` | `FALSE` | As in `lmer()` |
| `fit_mer` | `TRUE` | `FALSE` skips all reference fits and returns design and identifiability only |
| `dispformula` | `~1` | `~1` (pooled \(\sigma^2\)) or `~<group>` (per-group \(\sigma^2_j\), Gaussian only; adds a `glmmTMB` fit) |
| `...` | — | Passed to `lme4::lFormula()` and the reference fit |

### 1.3 Returned fields

| Field(s) | Content |
|----------|---------|
| `y`, `weights`, `offset` | Response, prior weights (ones by default) and offset (zeros by default), aligned with the rows of `D` |
| `D` | \(n \times p_{\mathrm{re}}\) matrix; column \(k\) is each observation's loading on random coefficient \(k\) |
| `group`, `group_name` | Grouping factor (with `attr(group, "group_name")`) and its name |
| `groupef.names` | Random-coefficient names; equal to `colnames(D)` and `names(W)` |
| `W` | Named list of \(J \times p_k\) hyper-design matrices, one row per group |
| `popef.moderation` | Data frame of cross-level interactions (`interaction_col`, `moderator`, `random_slope`) |
| `lmer` or `glmer` | Reference fit (`NULL` if `fit_mer = FALSE`) |
| `varcorr`, `Psi`, `dispersion` | `VarCorr()` table, named vector of \(\hat\tau^2_k\), residual \(\hat\sigma^2\) |
| `glmmTMB_fit` | Per-group-dispersion reference fit, or `NULL` |
| `groupef.rank`, `groupef.estimable`, `groupef.glm_check` | Level 1 diagnostics (Section 2.4) |
| `popef.rank`, `popef.deficient`, `popef.rank_ok` | Level 2 diagnostics (Section 2.4) |
| `formula`, `family`, `call`, `terms`, `data`, `dispformula` | Inputs, recorded as in `lm()`/`glm()` |

### 1.4 `check_identifiability()`

`check_identifiability(y, D, group, W, family = gaussian(), group_name, verbose = FALSE)`
runs the identifiability checks on any `D`/`group`/`W` triple.
`model_setup()` calls it internally, but it is exported so that hand-built
designs can be checked too (Section 2.5). `group` must be a factor,
`colnames(D)` must equal `names(W)`, and each `W[[k]]` must have one row per
level of `group`, in level order. With `verbose = TRUE` it reports progress
through `message()`.

## 2. Example: Big Word Club

The examples in Part 4 use the `big_word_club` data from **bayesrules**:
students (observations) nested in schools (groups), with a vocabulary test
score as the response.

```{r data, eval = has_br}
data(big_word_club, package = "bayesrules")
dat <- big_word_club
dat$school_id <- factor(dat$school_id)
dat <- subset(
  dat,
  !is.na(score_ppvt) &
    !is.na(invalid_ppvt) & invalid_ppvt == 0L &
    complete.cases(dat[, c(
      "score_ppvt", "distracted_a1", "distracted_ppvt",
      "private_school", "title1", "free_reduced_lunch", "school_id"
    )])
)

form_lmer <- score_ppvt ~
  private_school + title1 + free_reduced_lunch +
  distracted_ppvt + distracted_a1 +
  free_reduced_lunch:distracted_a1 +
  (1 + distracted_ppvt + distracted_a1 || school_id)
```

The model has three random coefficients per school: an intercept and two
slopes. The fixed part has three school-level covariates, two population
mean slopes and one cross-level interaction.

```{r design, eval = has_br}
design <- model_setup(form_lmer, data = dat)
design
```

### 2.1 The sampler design

```{r design-D, eval = has_br}
dim(design$D)
colnames(design$D)
design$group_name
nlevels(design$group)
```

### 2.2 The hyper-design matrices

```{r design-W, eval = has_br}
lapply(design$W, colnames)
design$popef.moderation
```

These follow the table in Section 1.1. The random intercept is regressed on
all three school-level covariates. The `distracted_a1` slope has its
population mean plus the moderator `free_reduced_lunch`. The
`distracted_ppvt` slope has only its population mean.

### 2.3 The reference fit

```{r design-ref, eval = has_br}
class(design$lmer)
design$Psi
design$dispersion
```

`Psi` holds the classical between-school variance \(\hat\tau^2_k\) of each
random coefficient, and `dispersion` holds the residual \(\hat\sigma^2\).
Set `fit_mer = FALSE` to skip the reference fits:

```{r design-only, eval = has_br}
design_only <- model_setup(form_lmer, data = dat, fit_mer = FALSE)
is.null(design_only$lmer)
```

### 2.4 Identifiability at two levels

**Level 1 (within each group).** Can the coefficient vector \(b_j\) be
estimated from group \(j\)'s data alone?

- `groupef.rank`: `TRUE` when the rows of `D` for group \(j\) have full
  column rank.
- `groupef.estimable`: full rank **and** a family-specific check passes.
  For Gaussian models the group needs more observations than coefficients
  (\(n_j > p_{\mathrm{re}}\)) so its residual variance is estimable. For
  binomial, Poisson and Gamma models a classical `glm()` is fitted to the
  group and must give finite estimates. This catches, for example, complete
  separation or an all-zero count response.
- `groupef.glm_check`: a data frame with a short note for each group that
  failed the family-specific check.

**Level 2 (across groups).** For each random coefficient \(k\),
`popef.rank[k]` is `TRUE` when `W[[k]]`, restricted to the Level 1
estimable groups, has full column rank. `popef.rank_ok` is `TRUE` when this
holds for every \(k\).

```{r ident, eval = has_br}
table(design$groupef.rank)
table(design$groupef.estimable)
design$popef.rank
design$popef.rank_ok
```

Groups that fail Level 1 stay in the reference fit. `Prior_Setup_GLMM()`
needs `popef.rank_ok = TRUE` to derive default hyperpriors. The Part 4
examples keep only full-rank groups before calibrating priors:

```{r full-rank, eval = has_br}
full_rank <- names(design$groupef.rank)[design$groupef.rank]
dat_fr <- subset(dat, school_id %in% full_rank)
dat_fr$school_id <- droplevels(dat_fr$school_id)

design_fr <- model_setup(form_lmer, data = dat_fr, fit_mer = FALSE)
all(design_fr$groupef.rank)
nlevels(design_fr$group)
```

### 2.5 Calling `check_identifiability()` directly

```{r ident-direct}
set.seed(1)
J <- 8L; n_j <- 6L
group <- factor(rep(paste0("g", seq_len(J)), each = n_j))
x1 <- rnorm(J * n_j)
D <- cbind("(Intercept)" = 1, x1 = x1)
y <- 1 + x1 + rnorm(J * n_j)
W <- list(
  "(Intercept)" = matrix(1, J, 1, dimnames = list(NULL, "(Intercept)")),
  x1            = matrix(1, J, 1, dimnames = list(NULL, "(Intercept)"))
)

ident <- check_identifiability(
  y = y, D = D, group = group, W = W,
  family = gaussian(), group_name = "group", verbose = TRUE
)
ident$popef.rank_ok
```

Now make one group's covariate constant. Its slope can no longer be
estimated from that group alone, so it fails Level 1:

```{r ident-direct-fail}
D_bad <- D
D_bad[group == "g3", "x1"] <- 0.5
ident_bad <- check_identifiability(
  y = y, D = D_bad, group = group, W = W, group_name = "group"
)
ident_bad$groupef.rank
```

## 3. Lineage: what runs inside `model_setup()`

### 3.1 Call tree

```
model_setup(formula, data, family, ..., fit_mer, dispformula)
 │
 ├─ .lmebayes_normalize_family()              family string/function → family object
 │
 ├─ extract_re_hyper_matrices()               ── STAGE A: sampler design
 │   ├─ is_single_factor_model()              exactly one grouping factor?
 │   ├─ .lmebayes_validate_uncorrelated_re_formula()
 │   │   └─ is_fixed_effects_only()           reject (1 || g) and correlated (x | g)
 │   ├─ lme4::lFormula()                      model frame, X, reTrms
 │   ├─ classify_lme4_fixed_columns()         split X into level-1 / level-2 columns
 │   ├─ classify_crosslevel_re_moderation()   level-1 columns → mean slopes,
 │   │                                        cross-level moderators, or disallowed
 │   ├─ get_lme4_components()                 X, sparse Z and its row/column maps
 │   │   └─ .lme4_Z_random_column_map(), .lme4_Z_random_row_map(),
 │   │      .lme4_label_Z_random_sparse()
 │   ├─ extract_lme4_fixed_group_matrix()     one row per group: level-2 X  → W
 │   │   └─ classify_lme4_fixed_columns()
 │   ├─ extract_re_Z_obs()                    sparse Z → dense n x p_re matrix D
 │   └─ .lmebayes_weights_offset_from_frame() weights/offset aligned with y
 │       └─ .lmebayes_normalize_weights(), .lmebayes_normalize_offset()
 │
 ├─ .lmebayes_dispformula_kind()              "pooled" (~1) or "group" (~<group>)
 │
 ├─ [fit_mer = TRUE]                          ── STAGE B: reference fits
 │   ├─ .lmebayes_mer_optional_args()         forward only supplied args
 │   ├─ lme4::lmer() | lme4::glmer()          always (backs design$lmer / $glmer)
 │   ├─ lme4::isSingular()                    message() if singular
 │   ├─ extract_mer_variance_components()     VarCorr → Psi, dispersion
 │   └─ [gaussian & dispformula = ~<group>]
 │       └─ .lmebayes_fit_glmmtmb_reference() glmmTMB(..., dispformula)
 │
 └─ check_identifiability()                   ── STAGE C: diagnostics
     ├─ Level 1 rank:   Matrix::rankMatrix() on each D_j
     ├─ Level 1 estimability:
     │   └─ .lmebayes_block_glm_estimable()
     │       └─ .lmebayes_glm_estimable_precheck()   quick family-specific rejects
     │          then stats::glm() per group           (binomial/poisson/Gamma)
     └─ Level 2 rank:   Matrix::rankMatrix() on W[[k]][estimable groups, ]
```

### 3.2 Helper reference

**Stage A — design extraction** (`R/lme4_design_utilities.R` unless noted)

| Helper | Role |
|--------|------|
| `extract_re_hyper_matrices()` | Orchestrates Stage A; returns the bare `"model_setup"` list (`y`, `D`, `group`, `W`, `popef.moderation`, `weights`, `offset`) |
| `is_single_factor_model()` | `TRUE` when `lFormula()` finds exactly one grouping factor |
| `is_fixed_effects_only()` | `TRUE` when there are no random-effect bars |
| `.lmebayes_validate_uncorrelated_re_formula()` | Stops on `(1 || g)` and on correlated multi-coefficient terms |
| `classify_lme4_fixed_columns()` | Splits the fixed-effect matrix into columns that are constant within groups (level 2) and columns that vary within groups (level 1) |
| `classify_crosslevel_re_moderation()` | Sorts level-1 columns into population mean slopes, `moderator:slope` interactions (`popef.moderation`) and disallowed columns |
| `get_lme4_components()` | Wraps `lFormula()`; returns `X`, the sparse random-effects `Z`, group factors and labelled row/column maps of `Z` |
| `.lme4_Z_random_column_map()`, `.lme4_Z_random_row_map()`, `.lme4_label_Z_random_sparse()` | Label the columns of `lme4`'s stacked `Z` by group, coefficient and level |
| `extract_lme4_fixed_group_matrix()` | Collapses the level-2 columns of `X` to one row per group; the source of every `W[[k]]` |
| `extract_re_Z_obs()` | Collapses the stacked sparse `Z` (one column per group × coefficient) to the dense \(n \times p_{\mathrm{re}}\) matrix `D` |
| `.lmebayes_weights_offset_from_frame()` (`mer_design_support.R`) | Reads `model.weights()` / `model.offset()` from the model frame and normalizes them to length \(n\) |
| `.lmebayes_normalize_family()` (`mer_family.R`) | Accepts a family name, function or object and returns a family object |

**Stage B — reference fits**

| Helper | File | Role |
|--------|------|------|
| `.lmebayes_mer_optional_args()` | `model_setup.R` | Collects `start`/`subset`/`weights`/`na.action`/`offset`/`contrasts` only if supplied, so `lmer()` sees the same call as a direct user call |
| `extract_mer_variance_components()` | `lme4_design_utilities.R` | Maps `VarCorr()` rows to `Psi` (named by `groupef.names`) and the residual variance |
| `.lmebayes_dispformula_kind()` | `glmmtmb_reference_helpers.R` | Validates `dispformula`; returns `"pooled"` or `"group"` |
| `.lmebayes_fit_glmmtmb_reference()` | `glmmtmb_reference_helpers.R` | Fits `glmmTMB(formula, dispformula = ~<group>)`; stored as `design$glmmTMB_fit` and reused by `Prior_Setup_GLMM()` |
| `.lmebayes_mer_convergence_issues()` | `model_setup.R` | Not called here; returns convergence problems of a `merMod` for `Prior_Setup_GLMM()` (S02) |
| `.lmebayes_stop_if_nondefault_weights_offset()` | `model_setup.R` | Not called here; a guard sampler engines call before ignoring non-default `weights`/`offset` |

**Stage C — identifiability** (`R/check_identifiability.R`)

| Helper | Role |
|--------|------|
| `check_identifiability()` | Exported; runs Level 1 rank, Level 1 estimability and Level 2 rank in that order |
| `.lmebayes_block_glm_estimable()` | Level 1 estimability for each full-rank group: an \(n_j > p\) check for Gaussian models, a classical `glm()` fit for binomial/Poisson/Gamma |
| `.lmebayes_glm_estimable_precheck()` | Rejects before fitting: a single outcome level (binomial), an all-zero response (Poisson), non-positive values or no residual df (Gamma) |

### 3.3 Exploring the internals

The non-dot design utilities are documented internally and can be called
with `:::`. For example, to see how the fixed columns were classified:

```{r internals, eval = has_br}
fr <- lme4::lFormula(form_lmer, data = dat_fr)
cls <- glmbayesCore:::classify_lme4_fixed_columns(fr$X, fr$reTrms$flist[[1]])
str(cls, max.level = 1)
```

## 4. Next

Core 20-S02 takes the full-rank design and uses `Prior_Setup_GLMM()` to
calibrate priors for both blocks.
